Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals

### 2016-02-19 manuscript/supplemental_material.Rnw
load libraries earlier and clearpage before sections

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[backend=biber,natbib=true,hyperref=true,style=nature,citestyle=numeric-comp,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage{float}
\floatstyle{plaintop}
\restylefloat{table}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\usepackage{cleveref}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\oddsidemargin 0.0in 
\textwidth 6.5in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
% make the afill smaller
\renewcommand\Affilfont{\itshape\small}
\renewcommand\cite[1]{\autocite{#1}}
\pagestyle{fancy}

@@

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE,cache.path="supplemental_cache/")
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -2, digits = 2)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ape")
library("ggtree")
library("topGO")
library("org.Hs.eg.db")

@@

<<source.to.latex,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R")))>>=
for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R"))) {
    source(a)
}
@

@@

\begin{table}
  \footnotesize
<<go.analysis.all.tissue,echo=FALSE,results="asis">>=
go.analysis.wide <- 
    data.table(dcast(data=go.results,
                     formula=GO.ID+type+Term+Ontology~species,
                     value.var="FDR"))
go.analysis.wide <- 
    go.analysis.wide[,`n(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
                                         function(x){sum(as.numeric(as.character(x)) <= 0.05)})]
go.analysis.wide <- 
    go.analysis.wide[,`max(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
                                           function(x){if (!any(is.finite(x) & x<=0.05)) {
                                                           return(NA)
                                                       } else {
                                                           max(x[is.finite(x) & x<=0.05],na.rm=TRUE)
                                                       }})]
go.analysis.wide.table <- 
    go.analysis.wide[!is.na(`n(FDR ≤ 0.05)`) & `n(FDR ≤ 0.05)` >= 12 &
                         grepl("Top 1% No Housekeeping",type),
                     list(GO.ID,Term,Ontology,`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
go.analysis.wide.table <- 
    go.analysis.wide.table[order(Ontology,-`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
setnames(go.analysis.wide.table,"GO.ID","GO ID")

print(xtable(go.analysis.wide.table,
             display=c("d","s","s","s","d","e"),
             align=  c("c","c","c","c","c","c")
             ),
      math.style.exponents=TRUE,
      include.rownames=FALSE,
      booktabs=TRUE,
      floating=FALSE
      )

@ 
\caption{GO terms with significant enrichment ($\mathrm{FDR}\le0.05$)
  in 12 or more of the species analyzed from genes in the top 1\% of
  expression. $n(\mathrm{FDR}\le0.05)$ is the number of species with
  significant enrichment ($\mathrm{FDR}\le 0.05$) in the GO term
  listed. $max(\mathrm{FDR}\le0.05)$ is the maximum FDR value in all
  species with an FDR of less than $0.05$.}
\label{tab:go_terms_all_species}
\end{table}

@@

\begin{table}
  \footnotesize
<<rodentia.lineage.specific,echo=FALSE,results="asis">>=
rod.all.lin.spec <- placenta.lineage.specific.table("rod.all")
setkey(rod.all.lin.spec,"Human Symbol")
@ 
\caption{Genes whose expression changed in the Rodentia
  order as compared to all other analyzed species.}
\label{tab:rodentia_lineage_specific}
\end{table}

\begin{table}
  \footnotesize
<<lau.lineage.specific,echo=FALSE,results="asis">>=
lau.all.lin.spec <- placenta.lineage.specific.table("lau.all")
setkey(lau.all.lin.spec,"Human Symbol")
@ 
\caption{Genes whose expression changed in the Laurasiatheria
  superorder as compared to all other analyzed species.}
\label{tab:laurasiatheria_lineage_specific}
\end{table}

\begin{table}
  \footnotesize
<<eua.lau.lineage.specific,echo=FALSE,results="asis">>=
eua.lau.lin.spec <- placenta.lineage.specific.table("eua.lau")
setkey(eua.lau.lin.spec,"Human Symbol")
@ 
\caption{Genes whose expression changed between the
  Euarchontoglires and Laurasiatherian superorder.}
\label{tab:laurasiatheria_euarchontoglires_lineage_specific}
\end{table}

<<top.expressed.genes,echo=FALSE,results="asis">>=
for (this.species in names(species.names)) {
    this.caption <- paste0("\\textit{",
                           capfirst(this.species),"} Placental Gene Expression of all genes with FPKM $\\ge 50$")
    cat(paste0("\\subsection{",this.caption,"}\n\n"))
    print(xtable(genes.expression.table(species.name=this.species),
                 caption=this.caption,
                 label=paste0("tab:all_",gsub(" ","_",this.species),"_placenta")
                 ),
          booktabs=TRUE,
          tabular.environment="longtable",
          floating=FALSE,
          caption.placement="top",
          sanitize.text.function=genes.table.sanitize,
          size="\\tiny"
          )
}
@

@@

>>>>

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[backend=biber,natbib=true,hyperref=true,style=nature,citestyle=numeric-comp,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage{float}
\floatstyle{plaintop}
\restylefloat{table}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\usepackage{cleveref}
\usepackage{xspace}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\oddsidemargin 0.0in 
\textwidth 6.5in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
% make the afill smaller
\renewcommand\Affilfont{\itshape\small}
\renewcommand\cite[1]{\autocite{#1}}
\pagestyle{fancy}

@@

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE,cache.path="supplemental_cache/")
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -2, digits = 2)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ape")
library("ggtree")
library("topGO")
library("org.Hs.eg.db")
library("magrittr")
library("tidyr")

@@

<<source.to.latex,echo=FALSE,message=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R")))>>=
for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R"))) {
    source(a)
}
@

@@

<<go.analysis.all.tissue,echo=FALSE,results="asis">>=
go.analysis.wide <- 
    data.table(dcast(data=go.results,
                     formula=GO.ID+type+Term+Ontology~species,
                     value.var="FDR"))
go.analysis.wide <- 
    go.analysis.wide[,`n(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
                                         function(x){sum(as.numeric(as.character(x)) <= 0.05)})]
go.analysis.wide <- 
    go.analysis.wide[,`max(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
                                           function(x){if (!any(is.finite(x) & x<=0.05)) {
                                                           return(NA)
                                                       } else {
                                                           max(x[is.finite(x) & x<=0.05],na.rm=TRUE)
                                                       }})]
go.analysis.wide.table <- 
    go.analysis.wide[!is.na(`n(FDR ≤ 0.05)`) & `n(FDR ≤ 0.05)` >= 12 &
                         grepl("Top 1% No Housekeeping",type),
                     list(GO.ID,Term,Ontology,`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
go.analysis.wide.table <- 
    go.analysis.wide.table[order(Ontology,-`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
setnames(go.analysis.wide.table,"GO.ID","GO ID")

print(xtable(go.analysis.wide.table,
             display=c("d","s","s","s","d","e"),
             align=  c("c","c","c","c","c","c"),
             caption="GO terms with significant enrichment ($\\mathrm{FDR}\\le0.05$)
  in 12 or more of the species analyzed from genes in the top 1\\% of
  expression. $n(\\mathrm{FDR}\\le0.05)$ is the number of species with
  significant enrichment ($\\mathrm{FDR}\\le 0.05$) in the GO term
  listed. $max(\\mathrm{FDR}\\le0.05)$ is the maximum FDR value in all
  species with an FDR of less than $0.05$.",
  label="tab:go_terms_all_species"
  ),
  math.style.exponents=TRUE,
  include.rownames=FALSE,
  booktabs=TRUE,
  floating=FALSE,
  tabular.environment="longtable",
  caption.placement="top",
  size="\\footnotesize"
  )

@@

@@

\begin{table}[hb]
  \footnotesize
<<rodentia.lineage.specific,echo=FALSE,results="asis">>=
rod.all.lin.spec <- placenta.lineage.specific.table("rod.all")
setkey(rod.all.lin.spec,"Human Symbol")
@ 
\caption{Genes whose expression changed in the Rodentia
  order as compared to all other analyzed species.}
\label{tab:rodentia_lineage_specific}
\end{table}

\begin{table}[hb]
  \footnotesize
<<lau.lineage.specific,echo=FALSE,results="asis">>=
lau.all.lin.spec <- placenta.lineage.specific.table("lau.all")
setkey(lau.all.lin.spec,"Human Symbol")
@ 
\caption{Genes whose expression changed in the Laurasiatheria
  superorder as compared to all other analyzed species.}
\label{tab:laurasiatheria_lineage_specific}
\end{table}

\begin{table}[hb]
  \footnotesize
<<eua.lau.lineage.specific,echo=FALSE,results="asis">>=
eua.lau.lin.spec <- placenta.lineage.specific.table("eua.lau")
setkey(eua.lau.lin.spec,"Human Symbol")
@ 
\caption{Genes whose expression changed between the
  Euarchontoglires and Laurasiatherian superorder.}
\label{tab:laurasiatheria_euarchontoglires_lineage_specific}
\end{table}

<<top.expressed.genes,echo=FALSE,results="asis">>=
for (this.species in names(species.names)) {
    this.caption <- paste0("\\textit{",
                           capfirst(this.species),"} Placental Gene Expression of all genes with FPKM $\\ge 50$")
    cat(paste0("\\subsection{\\textit{",capfirst(this.species),"} Placental Gene Expression}\n\n"))
    print(xtable(genes.expression.table(species.name=this.species),
                 caption=this.caption,
                 label=paste0("tab:all_",gsub(" ","_",this.species),"_placenta")
                 ),
          booktabs=TRUE,
          tabular.environment="longtable",
          floating=FALSE,
          caption.placement="top",
          sanitize.text.function=genes.table.sanitize,
          size="\\tiny"
          )
}
@
